---
format:
    gfm: default
---

# A Fantasy Football Points Projector

Who should I start this week in fantasy football? Fantasy football is a multi-billion-dollar industry with almost every major publication leading their weekend coverage with "Start Sit Advice for week X." In the platform you use there is a proprietary model that outputs a single forecast for the upcoming week. However, we all intuitively understand that Football has a lot of uncertainty. Take week 2 of last years NFL season if you had both Joe Burrow and Russell Wilson you may have debated about who to start. Joe Burrow has a much higher floor and boom potential, but Russell Wilson was playing against, what we would come to know, as one of the worst defenses in by DVOA in quite awhile. While less likely Wilson had a variety of scenarios where he could put up a vintage Russell Wilson performance. Projections heading into the game were fairly similar. Wilson was slightly ahead (14.5 vs. 13.4 expected points). That week, Wilson went for 30.3 real points against Burrow's 7.0, a 23.3-point swing. 

So who should have this person started going into the game? Since we were little we were probably told to think through the worst case outcomes before making a decision. Instead of using rules of thumb or preconceptions about players or matchups heading into the game, what if we could incorporate the uncertainty in our forecasts and a set of the worst case decisions to make more informed roster decisions?  

Throughout the last four years of playing fantasy football, I have found myself wanting a bit more from the predictions that my platform produces every week. Part of it is that it only produces a single point estimate and some, what I presume, MCMC simulations of the floor, ceiling, and mean outcome. Personally, I would like to see all this as I agonize over whether I should start or sit a player. So I decided to build a fully Bayesian pipeline for forecasting and making roster decisions using a similar setup to [this PyMC Labs blog post](https://www.pymc-labs.com/blog-posts/probabilistic-forecasting-optimization-under-uncertainty). The approach only works if the predictive intervals are honest, so most of the effort below goes into checking that they are.


```{r}
#| warning: false
#| message: false
#| echo: false
library(ggrepel)
library(MetBrewer)
library(tinytable)
library(jsonlite)
suppressPackageStartupMessages(library(tidyverse))

# one JSON per walk-forward fold, written by `python src/ff.py backtest`

files = list.files('results/backtest/calib', pattern = '\\.json$', full.names = TRUE)



folds_df = map(files, read_json) |>
  map(\(r) as_tibble(keep(r, \(v) length(v) == 1 && !is.list(v)))) |>
  list_rbind() |>
  mutate(model = ifelse(model == 'het_gp_qb_vegas', 'QB Model', 'Skill Model'))



coverage_summary = folds_df |>
    summarise(across(c(cov50, cov80, cov90), \(x) weighted.mean(x, w = n)), .by = model) |>
    pivot_longer(
        cols = starts_with("cov"),
        names_to = "nominal",
        names_pattern = "cov(\\d+)",
        values_to = "actual"
    ) |>
    mutate(nominal = as.numeric(nominal), actual = actual * 100)



label_data = coverage_summary |> filter(nominal == max(nominal))

p1 = ggplot(coverage_summary, aes(nominal, actual, color = model, group = model)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60", linewidth = 0.6) +
    geom_line(linewidth = 0.9, alpha = 0.6) +
    geom_point(size = 2.8, alpha = 0.6) +
    geom_label_repel(
        data = label_data, aes(label = model),
        size = 3.0, fontface = "bold", show.legend = FALSE
    ) +
    scale_color_met_d(name = 'Lakota') +
    scale_x_continuous(breaks = c(50, 80, 90), limits = c(0, 105), labels = \(x) paste0(x, "%")) +
    scale_y_continuous(breaks = seq(0, 100, 25), limits = c(0, 100), labels = \(x) paste0(x, "%")) +
    coord_fixed() +
    labs(
        x = "Nominal interval level",
        y = "Empirical coverage",
        title = "Interval calibration: empirical vs. nominal coverage",
        caption = "Dashed grey diagonal = perfect calibration. 12 walk-forward folds, 2023-2025."
    ) +
    AllenMisc::theme_allen_minimal() +
    theme(legend.position = 'none')

p1
```

Across 12 walk-forward folds (weeks 4, 8, 12 and 16 of 2023-2025), the 50/80/90% intervals hold the real score 58/85/92% of the time for RB/WR/TE and 49/79/89% for QB. That is close enough to trust a chance-constrained call, with one known problem covered in [Results](#results): for low projections the intervals are too wide.

## Setup

Dependencies are managed with [`uv`](https://docs.astral.sh/uv/pip/packages/#installing-a-package). To run this project, you need Python 3.14 installed. To replicate this project, just run

```
uv sync
```

## Weekly use

Everything runs through one command-line entry point, from the project root:

```bash
# 1. refresh the data through last week (needs network)
uv run python src/ff.py build-data
# 2. refresh team form
uv run python src/ff.py build-team-form
# 3. fit both models and forecast the coming week -> forecasts/2026_wk05/
uv run python src/ff.py forecast 2026 5
# 4. choose a lineup
# your roster can also we a csv file
uv run python src/ff.py decide forecasts/2026_wk05 --roster my_roster.txt --rule maximize_expected
uv run python src/ff.py decide forecasts/2026_wk05 --roster my_roster.txt --rule cvar_averse --lam 2
uv run python src/ff.py decide forecasts/2026_wk05 --roster my_roster.txt --rule head_to_head --opponent their_roster.csv
```

`decide` also reports the runner-up lineup and flags a **near tie** when the gap between them is within simulation noise, so a coin-flip start/sit call doesn't look like a confident one. `src/README.md` has the details: every command, the backtest and league-simulation tools, and what changed from the research code.

## Testing

```
uv run pytest src/tests -q
```

The fast tests build everything against a small synthetic league; tests that need the real parquet files skip themselves if the files are missing. Slow model fits are opt-in:

```
RUN_SLOW_TESTS=1 uv run pytest src/tests -q
```

## Project Structure

```
fantasy-predictor/
├── src/
│   ├── ff.py                     # command-line entry point
│   ├── ffpred/
│   │   ├── config.py             # paths, roster shape, model configs, sampler settings
│   │   ├── data_build.py         # nflverse -> processed-data/ff-processed.parquet
│   │   ├── team_form.py          # off/def team-strength state-space model -> team_form.parquet
│   │   ├── data.py               # loading, Vegas features, as-of team-form join, training windows
│   │   ├── upcoming.py           # model rows for a week that hasn't been played yet
│   │   ├── model.py              # the het_gp models: build, fit, predict, diagnostics
│   │   ├── forecast.py           # both models -> one week's joint predictive draws
│   │   ├── decision_engine.py    # lineup rules
│   │   ├── backtest.py           # walk-forward folds, one JSON per fold
│   │   ├── league.py             # league simulation of the decision rules
│   │   └── scoring.py            # CRPS / RMSE / coverage / tail calibration
│   ├── scratch.py                # model experiments
│   └── tests/
├── processed-data/
│   ├── ff-processed.parquet
│   └── team_form.parquet
├── results/
│   └── backtest/calib/           # the walk-forward folds behind the plots above
├── README.qmd
├── README.md
├── pyproject.toml
└── uv.lock
```

The research code that got the project here (the earlier model families, ablations and the original backtest harness) is in the git history; see [How the model evolved](#how-the-model-evolved).

## Data
All data are derived from [`nflreadr`](https://nflreadr.nflverse.com/articles/index.html). To train this model, I use data from 2013, when snaps are recorded, to 2025. 

- Fantasy Football Opportunity: A precomputed expected fantasy points dataset. I use: 
    - Total Fantasy Points: outcome variable of interest
    - Pass attempts, rush attempts, receptions, and yardage to compute efficiency and usage statistics
    - Fantasy points allowed: A rolling average of fantasy points over expectation allowed by the opponent to a position group
- Schedules: a dataset containing schedule information. I use: 
    - Spread line: The spread line for the game. 
    - Implied team total
    - Surface: collapsed to a binary variable where 1 indicates that the game is being played on grass
    - Roof: collapsed to a binary variable where 1 indicates whether the game was played indoors
    - Temperature: Temperature at the stadium
    - Wind: Speed of the wind in miles per hour. 
    - Division Game: a binary variable where 1 indicates whether the game is against a divisional opponent
    - Home team: collapsed to a binary variable where 1 indicates that the player is on the home team. 
    - Era: a binary indicator where 1 indicates the game occurs after the 2018 rule change 
- Snap counts: a dataset containing snap counts for each player: 
    - Offensive percent: percent of snaps taken on offense
    - Special teams percent: percent of special teams snaps taken.
- Roster: a dataset containing information on every NFL roster. I use: 
    - Rookie Year and Season to construct a player tenure variable.

- Vegas lines (from the schedules): converted to the player's side of the game.
    - Team spread: the spread from the player's team's point of view (positive = favoured)
    - Implied team total: half the game total plus half the team spread, i.e. how many points Vegas expects the player's team to score
- Team form: a dynamic offense/defense team-strength model (a state-space model on score margins, filtered with a Kalman filter) gives each team's form going into the week. Only weeks strictly before the game are used.

- Fantasy Football Player IDs I use: 
    - Pro Football Reference ID: merge key used to join snaps to main dataset

## Model Architecture

There are two models with the same structure: one for RB/WR/TE and a separate one for QBs, whose scoring and aging look quite different. Points are standardized within the training window, and each player's expected score for a week is built from:

- **A position intercept.**
- **A career aging curve.** A smooth bump over a player's tenure, with a shared width and a peak timing and height that vary by position: wide receivers can break out as rookies while tight ends tend to peak later. The QB model uses age instead of tenure and lets the curve rise and fall at different rates.
- **Player skill.** A career-long level for each player, partially pooled within his position.
- **Player form.** A Gaussian process over each player's seasons (Matérn-3/2 over tenure), so a player can be better or worse than his career level in a given year without that leaking into other years.
- **Game context.** Standardized usage and efficiency lags (target share, rush share, snap shares, yards per target, depth of target, ...), the opponent's recent points allowed to the position, weather, the player's side of the Vegas line and implied team total, binary game conditions, and both teams' form going into the week.

For RB/WR/TE the noise is heteroskedastic: each position has its own baseline spread, and a second Gaussian process lets the spread change over a career. The QB model uses a single noise scale. The inspiration for this idea came from [the Heteroskedastic Gaussian Process example from PyMC](https://www.pymc.io/projects/examples/en/latest/gaussian_processes/GP-Heteroskedastic.html). 

Fits use NUTS (via nutpie) with 4 chains of 1,000 draws after 1,000 tuning steps. The skill-position model takes about 3 minutes per fold and the QB model under a minute.

### How the model evolved

The first three candidates were chosen by expected log pointwise predictive density (ELPD) from a larger set. Each was a two-component mixture in which recent games missed and lagged snap share set the weight between a "limited role" and a "full-time role", so no player had to be dropped for low participation.

**ar_mod** was the baseline: flat position intercepts, a Gaussian random walk over tenure for latent skill and aging, and a player-season intercept for in-season form. The AR structure draws heavily on Dave Zach's [implementation in his spread model](https://github.com/dave-zack3/nfl_bayesian_rate_model/tree/main/src/models/spread), applied to opponent-defense strength.

**hs_version** gave each position its own intercept and tenure slope, and replaced the player-season intercept with a Hilbert-space Gaussian process over a player's season.

**team_mod** dropped the player-season intercepts and added the latent strength of both teams through a state-space model on rest-adjusted score margins, inspired by [Chris Fonnesbeck's World Cup model](https://www.pymc-labs.com/blog-posts/forecasting-the-2026-world-cup-group-stage).

Stacking, R2D2 shrinkage priors and BART were also tried. The current models came out of those experiments: the aging curve and per-player season GP replaced the random walk and season intercepts. After rethinking what the latent qualities of the model are I decided experimented and found that the Heteroskedastic GP tapped those latent quantities better. I found, through ablation testing, that the GP model with implied team totals and player side Vegas lines outperformed prior iterations of the model and was significantly faster. 
## Results

Each fold fits on the three seasons before the target week and forecasts that week's players; the baseline is each player's past average.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: summary-table



summary_stats = folds_df |>
    summarise(
        Folds = n(),
        Players = sum(n),
        RMSE = sqrt(weighted.mean(rmse^2, n)),
        Baseline = sqrt(weighted.mean(rmse_base^2, n)),
        MAE = weighted.mean(mae, n),
        CRPS = weighted.mean(crps, n),
        `Avg fit time (min)` = mean(fit_time) / 60,
        .by = model
    ) |>
    mutate(
        `RMSE % improvement vs baseline` = (Baseline - RMSE) / Baseline * 100,
        across(where(is.double), \(x) round(x, 2))
    ) |>
    select(model, RMSE, Baseline,`RMSE % improvement vs baseline` , CRPS, `Avg fit time (min)`) |> 
    rename(` ` = model) 

tt(summary_stats) |>
    theme_tinytable()

min_rmse =  min(summary_stats$`RMSE % improvement vs baseline`)
max_rmse = max(summary_stats$`RMSE % improvement vs baseline`)
```

The models beat the past-average baseline by `r min_rmse` for skill positions  and `r max_rmse` for QBs. This makes some sense if we know that the player is LaDanian Tomlinson in 2006 than we have a good idea of how well that player is going to play next week. However, the decision engine needs more than an accurate mean: a chance-constrained call that rests on a player clearing a threshold x% of the time is meaningless if the intervals are not calibrated. Overall coverage, in the plot at the top, is close to nominal. Broken down by how many points a player is projected for, it is less even:

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: tier-table

scalars = \(x) {
  as_tibble(keep(x, function(v) length(v) == 1 && !is.list(v)))}

tiers_df <- tibble(file = files) |>
   mutate(data = map(file, read_json)) |>
  mutate(
    model  = map_chr(data, "model"),
    season = map_int(data, "season"),
    week   = map_int(data, "week"),
    by_tier = map(data, "by_tier"), 
  ) |>
  select(-file, -data) |> 
  mutate(tier_data = map(by_tier, \(bt) map(bt, scalars) |> list_rbind())) |> 
  select(-by_tier) |>
  unnest(tier_data) |> 
  mutate(tier = factor(tier, levels = c('<6', '6-12', '12-18', '18+')),
         model = ifelse(model == 'het_gp_qb_vegas', 'QB Model', 'Skill Model'))


tiers_df

tier_summary = tiers_df|>
    summarise(
        Players = sum(n),
        across(c(mean_expected, mean_actual, below_p10, above_p90, pred_sd, resid_sd),
               \(x) weighted.mean(x, n)),
        .by = c(model, tier)
    ) |>
    arrange(desc(model), tier) |>
    mutate(
        across(c(below_p10, above_p90), \(x) 100 * x),
        across(where(is.double), \(x) round(x, 1)),
        model = ifelse(model == 'QB Model', model, 'Skill position')
    ) |>
    rename(
        Model = model, `Projection tier` = tier,
        `Mean projected` = mean_expected, `Mean actual` = mean_actual,
        `% below p10` = below_p10, `% above p90` = above_p90,
        `Predicted sd` = pred_sd, `Actual sd of misses` = resid_sd
    )

tt(tier_summary) |>
    theme_tinytable()
```

A calibrated model would put 10% of outcomes below the 10th percentile and 10% above the 90th in every tier, with the predicted and actual spread matching. Instead the model gives every player roughly the same spread (about 6 points for skill positions), while real spread grows with the projection: about 4 points for players projected under 6, about 8 for players projected 12-18. So the floor for low-projection players is too pessimistic (no outcomes fell below the 10th percentile), and the range for high-projection players is too narrow. Letting the spread scale with the projection is the main thing in progress; see the to-do list.

So how do we use this thing? Let's take the Yahoo Fantasy Expert's draft from 2025 as an imperfect example. We can't observe the week-to-week add-drops or waiver-wire activity, so we can't account for somebody adding Daniel Jones in the first week or two of the season or dropping Mike Evans when it was clear that his hamstring injury would prevent him from playing the rest of the year. I built a small simulator where we can build a round-robin schedule and have the lineups face each other. For demonstration purposes, I ran the decision engine on weeks 2-4 since we don't have to worry about bye weeks. I will focus on team 5 in week 2 because this is the first week where the decision engine had a different lineup than maximizing expected points.

Normally what I would do is just look at the players with the highest projected point totals and then maybe adjust who is starting versus who is sitting based on heuristics about matchup and potential game conditions. So my lineup would look, mostly like the table below 

```{r}
#| echo: false
#| warning: false
#| message: false
example_decisions = read_csv('results/decision_example.csv') |> 
    filter(same_as_baseline == FALSE) |> 
select(week, rule, lineup_type, player, realized_pts, predicted_pts = predicted_expected_pts, won) 


example_decisions |> 
    filter(lineup_type == 'baseline', week == 2) |> 
    select(player, predicted_pts, realized_pts, won) |> 
    mutate(across(ends_with('pts'), \(x) round(x,2))) |> 
    tt()
```


However, this is just one rule. The decision engine has a few more. Conditional Value at Risk (CVaR) trades some expected points for a more stable floor. First we pick a target, a fixed point threshold, which in this example is our opponent's projected total for the week, and an alpha, the fraction of worst-case scenarios we care about (20% here). To see what this is actually doing, I worked through a small example: ten representative weekly outcomes for a steady-floor player and a boom/bust player. For every scenario, we compute the shortfall below target, then average the shortfalls in just the worst 20% of scenarios. That average is the CVaR. The decision rule then maximizes expected points minus a penalty term, lambda, times that CVaR. Lambda controls how much of the tradeoff to make. At lambda=0, this is identical to just starting the highest-expected-points player. As lambda grows, the rule increasingly favors the player whose bad weeks are smaller in magnitude, even at the cost of a lower mean. In a practical sense, all we are doing is setting how much average performance you are willing to give up to make sure the floor doesn't collapse. 


```{r}
draws = tibble(
  player = rep(c("Player A (boom/bust)", "Player B (steady floor)"), each = 10),
  total  = c(2, 4, 6, 8, 10, 12, 22, 28, 32, 36,      # A
            10, 11, 12, 13, 14, 15, 16, 17, 18, 19)  # B
)

target = 12
alpha  = 0.20   # worst 20% of scenarios

cvar_tbl = draws |>
  mutate(shortfall = pmax(target - total, 0)) |>
  group_by(player) |>
  summarise(
    expected = mean(total),
    cvar     = shortfall |> sort(decreasing = TRUE) |> head(ceiling(n() * alpha)) |> mean()
  )


lambdas = seq(0, 1, by = 0.1)

objective_curve = cvar_tbl |>
  crossing(lambda = lambdas) |>
  mutate(objective = expected - lambda * cvar) |>
  select(player, lambda, objective) 



```


So our CVaR framework decided on this lineup. The main difference is that, due to the CVaR scoring, we ended up starting Joe Burrow over Russell Wilson, even though the model's own projections had Wilson slightly ahead (14.5 vs. 13.4 expected points). That week, Wilson went for 30.3 real points against Burrow's 7.0, a 23.3-point swing we gave up in exchange for a lower downside tail at QB. As it happened, the team won the matchup either way, so this particular week is a clean example of what the CVaR premium actually costs: real points given up for protection the lineup didn't end up needing. Meanwhile at RB, the same logic worked in the more typical direction, starting Tony Pollard over Tyrone Tracy, who has a higher ceiling but whose worst 20% of scenarios are meaningfully worse than Pollard's.


```{r}
#| echo: false
example_decisions |>  
    filter(week == 2,, lineup_type == 'rule') |>  
    filter(player %in% c('Joe Burrow', 'Tony Pollard')) |> 
    select(player, predicted_pts, realized_pts, won) |>
    relocate(player, .before = predicted_pts) |> 
    mutate(across(ends_with('pts'), \(x) round(x,2))) |> 
    tt()



```

## Decision Engine

The decision engine takes the joint predictive draws for a roster and picks a lineup under one of several rules:

- `maximize_expected`: the highest expected total. What 
- `chance_constrained`: the best expected total among lineups that clear a points floor in at least a given share of simulated weeks.
- `cvar_averse`: expected points minus a penalty on the average shortfall in the worst weeks (the CVaR example above).
- `head_to_head`: the lineup most likely to beat a specific opponent's lineup.
- `pareto_frontier`: the trade-off between expected points and downside risk, so you can see what each step of safety costs.

The structure (scenarios from the posterior predictive, then optimization over them) follows the [PyMC Labs post](https://www.pymc-labs.com/blog-posts/probabilistic-forecasting-optimization-under-uncertainty) and FICO's [stochastic energy-planning example](https://github.com/fico-xpress/xpress-community/blob/main/StochasticEnergyPlanning/stochastic_energy_planning.py) for Xpress. Small rosters are solved by enumerating every lineup; mixed-integer versions are available when the Xpress solver is installed.

## To Do

- [x] Build a more robust decision engine to add in Pareto Analysis similar to [the PyMC blog post](https://www.pymc-labs.com/blog-posts/probabilistic-forecasting-optimization-under-uncertainty)
- [x] Improve the organization of the repo so that we can use various scripts as modules and rewire the underlying scripts.
- [x] Report near ties so coin-flip start/sit calls are flagged as such.
- [ ] Let the predictive spread grow with the projection (fixes the tier calibration above). A first version improved CRPS by about 2% but pulled down projections for top players, because tying the spread to the mean biases the mean; the next version ties it to the player's trailing average instead.
- [ ] Compare lineups chosen with and without that change by realized points, not just forecast scores.
- [ ] Try out [Alchemize](https://github.com/pymc-labs/alchemize) to rebuild the underlying models in Rust to speed up sampling.
- [ ] Build a friendlier interface to use.
- [ ] Add a scheduler so I don't have to run this manually every week.
