---
format:
    gfm: default
---


# A Fantasy Football Points Projector 

Throughout the last four years of playing fantasy football, I have found myself wanting a bit more from the predictions that my platform produces every week. Part of it is that it only produces a single point estimate and some, what I presume, MCMC simulations of the floor, ceiling, and mean outcome. Personally, I would like to see all this as I agonize over whether I should start or sit a player. So I decided to build a fully Bayesian pipeline for forecasting and making roster decisions using a similar setup to [this PyMC Labs blog post](https://www.pymc-labs.com/blog-posts/probabilistic-forecasting-optimization-under-uncertainty). 

## Setup 

Dependencies are managed with [`uv`](https://docs.astral.sh/uv/pip/packages/#installing-a-package). To run this project, you need Python 3.14 installed. To replicate this project, just run 

```
uv sync
```

## Project Structure

```
fantasy-predictor/
├── src/                          # data pipeline, model definitions, backtesting
│   ├── data-cleaning.py
│   ├── ff-ar.py                  # ar_mod / hs_version / team_mod
│   ├── bart-mod.py
│   ├── estimate-latent-ability.py
│   ├── decision_engine.py
│   ├── league_validate.py
│   ├── calibrate_decisions.py
│   ├── forecast_roster.py
│   ├── mock_league.py
│   ├── adp_league.py
│   ├── backtest_harness.py       # walk-forward CRPS/RMSE/coverage backtest
│   ├── summarize_backtest.py
│   ├── sweep_summarize.py
│   └── run_*.sh                  # backtest / calibration / sweep drivers
├── scripts/
│   └── show_decision.py          # prints/writes the decision engine's actual lineup picks
├── eval_scripts/
│   └── eval_utils.py
├── processed-data/
│   ├── ff-processed.parquet
│   └── team_form.parquet
├── results/
│   ├── backtest_walkforward.csv
│   ├── decision_example.csv
│   └── league_validation.csv
├── README_files/
│   └── figure-commonmark/        # plot images rendered into README.md
├── README.qmd
├── README.md
├── pyproject.toml
├── uv.lock
├── skills-lock.json
└── .gitignore
```

## Data
All data are derived from [`nflreadr`](https://nflreadr.nflverse.com/articles/index.html). To train this model, I use data from 2013, when snaps are recorded, to 2025. 

- Fantasy Football Opportunity: A precomputed expected fantasy points dataset. I use: 
    - Total Fantasy Points: outcome variable of interest
    - Pass attempts, rush Attempts, receptions, and yardage to compute efficiency and usage statistics
    - Fantasy points allowed: A rolling average of fantasy points over expectation allowed by the opponent to a position group
- Schedules: a dataset containing schedule information. I use: 
    - Spread line: The spread line for the game 
    - Surface: collapsed to a binary variable where 1 indicates that the game is being played on grass
    - Roof: collapsed to a binary variable where 1 indicates whether the game was played indoors
    - Temperature: Temperature at the stadium
    - Wind: Speed of the wind in miles per hour. 
    - Division Game: a binary variable where 1 indicates whether the game is against a divisional opponent
    - Home team: collapsed to a binary variable where 1 indicates that the player is on the home team. 
    - Era: a binary indicator where 1 indicates the game occurs after the 2018 rule change 
- Snap counts: a dataset containing snap counts for each player: 
    - Offensive percent: percent of snaps taken on offense
    - Special teams percent: percent of special teams snaps taken.
- Roster: a dataset containing information on every NFL roster. I use: 
    - Rookie Year and Season to construct a player tenure variable.

- Fantasy Football Player IDs I use: 
    - Pro Football Reference ID: merge key used to join snaps to main dataset

## Model Architecture 

3 candidate models were eventually settled on using expected log pointwise predictive density (ELPD) against a lot of candidate models. Each of the three models shares some common components. Each of them have the same priors on continuous and binary covariates. Each model is a two-component mixture where a player's recent games-missed and lagged offensive snap share determine the mixtures weight between a "limited role" and a "full-time role." In the spirit of building a full fantasy prediction machine I didn't want to drop any players conditional on offensive participation. 

To get into the differences

**ar_mod**: This was effectively the baseline model. I use flat position intercepts, a Gaussian random-walk over the number of years a player has played in the NFL to measure latent skill and age curves, and a player season intercept to capture deviations in player form in that particular season. The AR structure draws heavily on Dave Zach's [implementation in his spread model](https://github.com/dave-zack3/nfl_bayesian_rate_model/tree/main/src/models/spread). I place this AR process on opponent-defense strength over the defense's week-to-week form. 

**hs_version**: This model has two main structural swaps. I build a varying effects model where each position gets its own position and its own slope based on standardized tenure. This *tries*, to capture the differing maturation curves of positional groups. In the past 5 years we consistently see Wide Receivers break out in their rookie year while Tight Ends tend to break out in their second or third year. The player season intercept is then replaced with a Hilbert-space Gaussian Process over a player's season to capture week-to-week booms or ramp up to injury. 

**team_mod**: This model differs by omitting the player season intercepts. I also modeled the latent ability of the player's team and the opponent via a state-space model on rest-adjusted score margins




## Results 
After running the ELPD comparisons, the results suggested that I should also try [model stacking](https://www.pymc.io/projects/examples/en/latest/diagnostics_and_criticism/model_averaging.html#stacking). I then ran a backtest on data from week 6 to week 15 for the 2013, 2019, 2023, and 2024 seasons. Looking at the RMSE, we see a modest improvement over just using a player's past mean. 


```{r}
#| echo: false
#| message: false 
#| warning: false
library(tinytable)
library(ggrepel)
suppressPackageStartupMessages(library(tidyverse))

backtests = read_csv('results/backtest_walkforward.csv')

summary_stats = backtests |> 
    summarise(
        across(c(rmse, mae, crps, fit_time), \(x) mean(x), .names = "{.col}_avg"), 
        .by = model
    ) |>
        arrange(rmse_avg) |> 
        mutate(
             rmse_avg_pct_improve = round((rmse_avg[model == "baseline"] - rmse_avg) / rmse_avg[model == "baseline"] * 100, 2),
            fit_time_avg  = fit_time_avg/60,
            model = case_match(model, 
            'team' ~ 'State-Space Model estimated team strength', 
            'stack' ~ 'Bayesian Model Stacking', 
            'ar' ~ 'AR(1) Process on Opp form', 
            'hs' ~ 'HSGP on player form')
        ) |> 
            filter_out(model == 'baseline') |> 
            pivot_longer(-model) |>
            mutate(
                name = case_match(name, 
                'crps_avg' ~ 'Mean CRPS', 
                'fit_time_avg' ~ 'Average Fit Time', 
                'mae_avg' ~ 'Mean MAE', 
                'rmse_avg' ~ 'Mean RMSE', 
                "rmse_avg_pct_improve" ~ "RMSE % Improvement vs Baseline"), 
                value = round(value, 3)
            )  |> 
                pivot_wider(
                    id_cols = model, names_from = name, values_from = value
                ) |> 
                    rename(` `= model) |> 
                    drop_na()


tt(summary_stats) |> 
    theme_tinytable()
```

However, part of the underlying motivation of this project was to build a decision engine where I can mitigate risky start-sit decisions. A chance-constrained call that rests on a player clearing a certain threshold x% of the time is meaningless if these intervals are not well calibrated. Each of these models clears this hurdle by about 2 percentage points across each specification. For example, if we set the threshold for making decisions at 50% and used the state-space model, the bands that this model can produce hold the true value about 52% of the time.


```{r}
#| echo: false
coverage_summary = backtests |>
    summarise(
        across(c(cov50, cov80, cov90), \(x) weighted.mean(x, w = n)),
        .by = model
    ) |>
    pivot_longer(
        cols = starts_with("cov"),
        names_to = "nominal",
        names_pattern = "cov(\\d+)",
        values_to = "actual"
    ) |>
    mutate(
        nominal = as.numeric(nominal),
        actual = actual * 100,
        model = case_match(model,
            "baseline" ~ "Baseline (point estimate)",
            "ar"       ~ "AR(1) Process on Opp form",
            "hs"       ~ "HSGP on player form",
            "team"     ~ "State-Space est team strength",
            "stack"    ~ "Bayesian Model Stacking"
        )
    )

label_data = coverage_summary |> filter(nominal == max(nominal))

ggplot(coverage_summary, aes(nominal, actual, color = model, group = model)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60", linewidth = 0.6) +
    geom_line(
        data = filter(coverage_summary, model == "Baseline (point estimate)"),
        linetype = "dashed", linewidth = 0.9
    ) +
    geom_line(
        data = filter(coverage_summary, model != "Baseline (point estimate)"),
        linewidth = 0.9, alpha = 0.5
    ) +
    geom_point(size = 2.8, alpha = 0.5, position = position_jitter(width = 0.05, seed = 1994)) +
    ggrepel::geom_label_repel(
        data = label_data, aes(label = model),
        size = 3.0, fontface = "bold", show.legend = FALSE
    ) +
    MetBrewer::scale_color_met_d(name = 'Lakota') +
    scale_x_continuous(breaks = c(50, 80, 90), limits = c(0, 105), labels = \(x) paste0(x, "%")) +
    scale_y_continuous(breaks = seq(0, 100, 25), limits = c(0, 100), labels = \(x) paste0(x, "%")) +
    coord_fixed() +
    labs(
        x = "Nominal interval level",
        y = "Empirical coverage",
        title = "Interval calibration: empirical vs. nominal coverage",
        caption = "Dashed grey diagonal = perfect calibration"
    ) +
    AllenMisc::theme_allen_minimal() + 
    theme(legend.position = 'none')
```

So how do we use this thing? Let's take the Yahoo Fantasy Expert's draft from 2025 as an imperfect example. We can't observe the week-to-week add-drops or waiver-wire activity, so we can't account for somebody adding Daniel Jones in the first week or two of the season or dropping Mike Evans when it was clear that his hamstring injury would prevent him from playing the rest of the year. I built a small simulator where we can build a round-robin schedule and have the lineups face each other. For demonstration purposes, I ran the decision engine on weeks 2-4 since we don't have to worry about bye weeks. I will focus on team 5 in week 2 because this is the first week where the decision engine had a different lineup than maximizing expected points.

Normally what I would do is just look at the players with the highest projected point totals and then maybe adjust who is starting versus who is sitting based on heuristics about matchup and potential game conditions. So my lineup would look, mostly like the table below 

```{r}
#| echo: false
#| warning: false
#| message: false
example_decisions = read_csv('results/decision_example.csv') |> 
    filter(same_as_baseline == FALSE) |> 
select(week, rule, lineup_type, player, realized_pts, predicted_pts = predicted_expected_pts, won) 


example_decisions |> 
    filter(lineup_type == 'baseline', week == 2) |> 
    select(player, predicted_pts, realized_pts, won) |> 
    mutate(across(ends_with('pts'), \(x) round(x,2))) |> 
    tt()
```


However, this is just one rule. The decision engine has a few more. Conditional Value at Risk (CVaR) trades some expected points for a more stable floor. First we pick a target, a fixed point threshold, which in this example is our opponent's projected total for the week, and an alpha, the fraction of worst-case scenarios we care about (20% here). To see what this is actually doing, I worked through a small example: ten representative weekly outcomes for a steady-floor player and a boom/bust player. For every scenario, we compute the shortfall below target, then average the shortfalls in just the worst 20% of scenarios. That average is the CVaR. The decision rule then maximizes expected points minus a penalty term, lambda, times that CVaR. Lambda controls how much of the tradeoff to make. At lambda=0, this is identical to just starting the highest-expected-points player. As lambda grows, the rule increasingly favors the player whose bad weeks are smaller in magnitude, even at the cost of a lower mean. In a practical sense, all we are doing is setting how much average performance am I willing to give up to make sure the floor doesn't collapse. 


```{r}
draws = tibble(
  player = rep(c("Player A (boom/bust)", "Player B (steady floor)"), each = 10),
  total  = c(2, 4, 6, 8, 10, 12, 22, 28, 32, 36,      # A
            10, 11, 12, 13, 14, 15, 16, 17, 18, 19)  # B
)

target = 12
alpha  = 0.20   # worst 20% of scenarios

cvar_tbl = draws |>
  mutate(shortfall = pmax(target - total, 0)) |>
  group_by(player) |>
  summarise(
    expected = mean(total),
    cvar     = shortfall |> sort(decreasing = TRUE) |> head(ceiling(n() * alpha)) |> mean()
  )


lambdas = seq(0, 1, by = 0.1)

objective_curve = cvar_tbl |>
  crossing(lambda = lambdas) |>
  mutate(objective = expected - lambda * cvar) |>
  select(player, lambda, objective) 



```


So our CVaR framework decided on this lineup. The main difference is that, due to the CVaR scoring, we ended up starting Joe Burrow over Russell Wilson, even though the model's own projections had Wilson slightly ahead (14.5 vs. 13.4 expected points). That week, Wilson went for 30.3 real points against Burrow's 7.0, a 23.3-point swing we gave up in exchange for a lower downside tail at QB. As it happened, the team won the matchup either way, so this particular week is a clean example of what the CVaR premium actually costs: real points given up for protection the lineup didn't end up needing. Meanwhile at RB, the same logic worked in the more typical direction, starting Tony Pollard over Tyrone Tracy, who has a higher ceiling but whose worst 20% of scenarios are meaningfully worse than Pollard's.


```{r}


example_decisions |>  
    filter(week == 2,, lineup_type == 'rule') |>  
    filter(player %in% c('Joe Burrow', 'Tony Pollard')) |> 
    select(player, predicted_pts, realized_pts, won) |>
    relocate(player, .before = predicted_pts) |> 
    mutate(across(ends_with('pts'), \(x) round(x,2))) |> 
    tt()



```



## To Do 

- [] Build a more robust decision engine to add in Pareto Analysis similar to [the PyMC blog post](https://www.pymc-labs.com/blog-posts/probabilistic-forecasting-optimization-under-uncertainty)
- [] Improve the organization of the repo so that we can use various scripts as modules and rewire the underlying scripts.
- [] Build a friendlier interface to use. 
- [] Add a scheduler so I don't have to run this manually everyweek 
